---
title: "Modelling and forecasting time series in R"
author: "**Dr Matt Ashby** | UCL Jill Dando Institute | `r format(Sys.Date(), '%B %Y')`"
output:
  xaringan::moon_reader:
    css: ["ucl-basic.css", "animate.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE)

library(ggtext)
library(gt)
library(lubridate)
library(tidyverse)

source(here::here("../helpers.R"))
```



class: animated, fadeIn

```{r chart-generic-time-series-1, fig.asp=0.56, fig.retina=4, out.width="100%"}
generic_time_series <- 0:23 %>% 
	enframe(name = "row", value = "date") %>% 
	mutate(
		trend = (row / 2) + 29,
		season = sin(seq(0, 4 * pi, length.out = n())) * 3,
		remainder = rnorm(n(), sd = 2.5)
	)

ggplot(generic_time_series) +
	geom_point(aes(date, trend + season + remainder), colour = "grey33", shape = 21, size = 2) +
	# geom_point(aes(date, trend), colour = ucl_colours_list["Bright Red"]) +
	# geom_line(aes(date, trend), colour = ucl_colours_list["Bright Red"], linetype = "43") +
	# geom_segment(aes(x = date, xend = date, y = trend, yend = trend + season), colour = ucl_colours_list["Bright Blue"]) +
	# geom_point(aes(date, trend + season), colour = ucl_colours_list["Bright Blue"]) +
	# geom_line(aes(date, trend + season), colour = ucl_colours_list["Bright Blue"], linetype = "43") +
	# geom_segment(aes(x = date, xend = date, y = trend + season, yend = trend + season + remainder), colour = ucl_colours_list["Orange"]) +
	# geom_point(aes(date, trend + season + remainder), colour = ucl_colours_list["Orange"]) +
	scale_x_continuous(n.breaks = 12) +
	scale_y_continuous(limits = c(0, NA), n.breaks = 10) +
	labs(
		x = "time →",
		y = "the thing you're measuring →"
	) +
	theme_minimal() +
	theme(
		axis.text = element_blank(),
		axis.ticks = element_line(colour = "grey33"),
		axis.title = element_markdown(),
		panel.grid = element_blank()
	)
```



---

class: animated, fadeIn

```{r chart-generic-time-series-2, fig.asp=0.56, fig.retina=4, out.width="100%"}
ggplot(generic_time_series) +
	geom_point(aes(date, trend + season + remainder), colour = "grey33", shape = 21, size = 2) +
	# geom_point(aes(date, trend), colour = ucl_colours_list["Bright Red"]) +
	geom_line(aes(date, trend), colour = ucl_colours_list["Bright Red"]) +
	# geom_segment(aes(x = date, xend = date, y = trend, yend = trend + season), colour = ucl_colours_list["Bright Blue"]) +
	# geom_point(aes(date, trend + season), colour = ucl_colours_list["Bright Blue"]) +
	# geom_line(aes(date, trend + season), colour = ucl_colours_list["Bright Blue"], linetype = "43") +
	# geom_segment(aes(x = date, xend = date, y = trend + season, yend = trend + season + remainder), colour = ucl_colours_list["Orange"]) +
	# geom_point(aes(date, trend + season + remainder), colour = ucl_colours_list["Orange"]) +
	annotate(
		geom = "text",
		x = max(generic_time_series$date), 
		y = 2, 
		label = "trend",
		hjust = 1,
		size = chart_elements$label_text_size * 2
	) +
	scale_x_continuous(n.breaks = 12) +
	scale_y_continuous(limits = c(0, NA), n.breaks = 10) +
	labs(
		x = "time →",
		y = "the thing you're measuring →"
	) +
	theme_minimal() +
	theme(
		axis.text = element_blank(),
		axis.ticks = element_line(colour = "grey33"),
		axis.title = element_markdown(),
		panel.grid = element_blank()
	)
```



---

class: animated, fadeIn

```{r chart-generic-time-series-3, fig.asp=0.56, fig.retina=4, out.width="100%"}
ggplot(generic_time_series) +
	geom_point(aes(date, trend + season + remainder), colour = "grey33", shape = 21, size = 2) +
	# geom_point(aes(date, trend), colour = ucl_colours_list["Bright Red"]) +
	geom_line(aes(date, trend), colour = ucl_colours_list["Bright Red"]) +
	geom_segment(aes(x = date, xend = date, y = trend, yend = trend + season), colour = ucl_colours_list["Bright Blue"]) +
	# geom_point(aes(date, trend + season), colour = ucl_colours_list["Bright Blue"]) +
	geom_line(aes(date, trend + season), colour = ucl_colours_list["Bright Blue"], linetype = "63") +
	# geom_segment(aes(x = date, xend = date, y = trend + season, yend = trend + season + remainder), colour = ucl_colours_list["Orange"]) +
	# geom_point(aes(date, trend + season + remainder), colour = ucl_colours_list["Orange"]) +
	annotate(
		geom = "text",
		x = max(generic_time_series$date), 
		y = 2, 
		label = "trend + seasonality",
		hjust = 1,
		size = chart_elements$label_text_size * 2
	) +
	scale_x_continuous(n.breaks = 12) +
	scale_y_continuous(limits = c(0, NA), n.breaks = 10) +
	labs(
		x = "time →",
		y = "the thing you're measuring →"
	) +
	theme_minimal() +
	theme(
		axis.text = element_blank(),
		axis.ticks = element_line(colour = "grey33"),
		axis.title = element_markdown(),
		panel.grid = element_blank()
	)
```



---

class: animated, fadeIn

```{r chart-generic-time-series-4, fig.asp=0.56, fig.retina=4, out.width="100%"}
ggplot(generic_time_series) +
	geom_point(aes(date, trend + season + remainder), colour = "grey33", shape = 21, size = 2) +
	# geom_point(aes(date, trend), colour = ucl_colours_list["Bright Red"]) +
	# geom_line(aes(date, trend), colour = ucl_colours_list["Bright Red"], linetype = "43") +
	# geom_segment(aes(x = date, xend = date, y = trend, yend = trend + season), colour = ucl_colours_list["Bright Blue"]) +
	# geom_point(aes(date, trend + season), colour = ucl_colours_list["Bright Blue"]) +
	geom_line(aes(date, trend + season), colour = ucl_colours_list["Bright Blue"], linetype = "63") +
	geom_segment(aes(x = date, xend = date, y = trend + season, yend = trend + season + remainder), colour = ucl_colours_list["Orange"]) +
	geom_point(aes(date, trend + season + remainder), colour = ucl_colours_list["Orange"]) +
	annotate(
		geom = "text",
		x = max(generic_time_series$date), 
		y = 2, 
		label = "trend + seasonality + everything else",
		hjust = 1,
		size = chart_elements$label_text_size * 2
	) +
	scale_x_continuous(n.breaks = 12) +
	scale_y_continuous(limits = c(0, NA), n.breaks = 10) +
	labs(
		x = "time →",
		y = "the thing you're measuring →"
	) +
	theme_minimal() +
	theme(
		axis.text = element_blank(),
		axis.ticks = element_line(colour = "grey33"),
		axis.title = element_markdown(),
		panel.grid = element_blank()
	)
```



---

class: big-text, animated, fadeIn

```{r chart-different-trends-1, fig.asp=0.56, fig.retina=4, out.width="100%"}
different_trends <- generic_time_series %>% 
	select(date, linear = trend) %>% 
	mutate(
		quadratic = linear + 2 + (date^2 / 20),
		something_else = linear - 2 - c(
			rep(0, times = 4), 
			abs(sin(seq(pi * 0.5, pi * 1.5, length.out = 16)) - 1) * 7.5, 
			rep(15, times = 4)
		)
	) %>% 
	pivot_longer(-date, names_to = "type", values_to = "trend")

different_trends %>% 
	filter(type == "linear") %>% 
	ggplot(aes(date, trend, colour = type)) + 
	geom_line() + 
	scale_x_continuous(n.breaks = 12) +
	scale_y_continuous(
		limits = c(0, max(different_trends$trend)), 
		n.breaks = 10
	) +
	scale_colour_manual(
		values = unname(ucl_colours_list[c("Bright Red", "Bright Blue", "Orange")])
	) +
	labs(
		x = "time →",
		y = "the thing you're measuring →"
	) +
	theme_minimal() +
	theme(
		axis.text = element_blank(),
		axis.ticks = element_line(colour = "grey33"),
		axis.title = element_markdown(),
		legend.position = "none",
		panel.grid = element_blank()
	)
```



---

class: big-text, animated, fadeIn

```{r chart-different-trends-2, fig.asp=0.56, fig.retina=4, out.width="100%"}
ggplot(different_trends, aes(date, trend, colour = type)) + 
	geom_line() + 
	annotate(
		geom = "text",
		x = min(different_trends$date), 
		y = max(different_trends$trend), 
		label = "trends can be non-linear",
		hjust = 0,
		size = chart_elements$label_text_size * 2,
		vjust = 1
	) +
	scale_x_continuous(n.breaks = 12) +
	scale_y_continuous(
		limits = c(0, max(different_trends$trend)), 
		n.breaks = 10
	) +
	scale_colour_manual(
		values = unname(ucl_colours_list[c("Bright Red", "Bright Blue", "Orange")])
	) +
	labs(
		x = "time →",
		y = "the thing you're measuring →"
	) +
	theme_minimal() +
	theme(
		axis.text = element_blank(),
		axis.ticks = element_line(colour = "grey33"),
		axis.title = element_markdown(),
		legend.position = "none",
		panel.grid = element_blank()
	)
```



---

class: big-text, animated, fadeIn

```{r chart-multiple-seasons-1, fig.asp=0.56, fig.retina=4, out.width="100%"}
multiple_seasons <- tibble(date = seq(0, (24 * 12) - 1)) %>% 
	mutate(
		trend = (date / 24) + 29,
		season1 = sin(seq(0, 6 * pi, length.out = n())) * 3,
		season2 = sin(seq(0, 6 * 8 * pi, length.out = n())) * 2,
		season3 = case_when(
			# First few days, since otherwise these match the subsequent conditions
			date <= 3 ~ 0,
			# Two days before/after peak day
			(date + 2) %% 72 == 0 | (date - 2) %% 72 == 0 ~ 1 + rpois(n(), lambda = 2),
			# Day before/after peak day
			(date + 1) %% 72 == 0 | (date - 1) %% 72 == 0 ~ 2.5 + rpois(n(), lambda = 2),
			# Peak day
			date %% 72 == 0 ~ 5 + rpois(n(), lambda = 2),
			# Every other day
			TRUE ~ 0
		)
	)

ggplot(multiple_seasons) + 
	geom_line(aes(date, trend), colour = "grey67") +
	geom_line(aes(date, trend + season1), colour = ucl_colours_list["Bright Blue"], linetype = "63") +
	# geom_line(aes(date, trend + season1 + season2), colour = ucl_colours_list["Bright Blue"], linetype = "32") +
	annotate(
		geom = "text",
		x = max(multiple_seasons$date), 
		y = 2, 
		label = "single seasonal pattern",
		hjust = 1,
		size = chart_elements$label_text_size * 2
	) +
	scale_x_continuous(n.breaks = 12) +
	scale_y_continuous(
		limits = c(
			0, 
			max(multiple_seasons$trend + multiple_seasons$season1 + multiple_seasons$season2 + multiple_seasons$season3)
		), 
		n.breaks = 10
	) +
	labs(
		x = "time →",
		y = "the thing you're measuring →"
	) +
	theme_minimal() +
	theme(
		axis.text = element_blank(),
		axis.ticks = element_line(colour = "grey33"),
		axis.title = element_markdown(),
		legend.position = "none",
		panel.grid = element_blank()
	)
```



---

class: big-text, animated, fadeIn

```{r chart-multiple-seasons-2, fig.asp=0.56, fig.retina=4, out.width="100%"}
ggplot(multiple_seasons) + 
	# geom_line(aes(date, 0), colour = "grey67") +
	# geom_line(aes(date, trend + season1), colour = ucl_colours_list["Bright Blue"], linetype = "63") +
	geom_line(aes(date, trend + season1 + season2 + season3), colour = ucl_colours_list["Bright Blue"], linetype = "32") +
	annotate(
		geom = "text",
		x = max(multiple_seasons$date), 
		y = 2, 
		label = "multiple seasonal patterns",
		hjust = 1,
		size = chart_elements$label_text_size * 2
	) +
	scale_x_continuous(n.breaks = 12) +
	scale_y_continuous(
		limits = c(
			0, 
			max(multiple_seasons$trend + multiple_seasons$season1 + multiple_seasons$season2 + multiple_seasons$season3)
		), 
		n.breaks = 10
	) +
	labs(
		x = "time →",
		y = "the thing you're measuring →"
	) +
	theme_minimal() +
	theme(
		axis.text = element_blank(),
		axis.ticks = element_line(colour = "grey33"),
		axis.title = element_markdown(),
		legend.position = "none",
		panel.grid = element_blank()
	)
```



---

class: big-text, middle, center, animated, fadeIn

time-series data typically violate<br>the assumptions of regression models



---

class: big-text, animated, fadeIn

```{r acf-norm, fig.asp=0.56, fig.retina=4, out.width="100%"}
acf(
	rnorm(nrow(multiple_seasons)), 
	main = "autocorrelation plot for random data",
	xlab = "lag",
	ylab = "correlation"
)
```



---

class: big-text, animated, fadeIn

```{r acf-ts, fig.asp=0.56, fig.retina=4, out.width="100%"}
multiple_seasons %>% 
	mutate(sum = trend + season1 + season2 + season3 + rnorm(n(), sd = 2.5)) %>% 
	pull(sum) %>% 
	acf(
		main = "autocorrelation plot for time-series data",
		xlab = "lag",
		ylab = "correlation"
	)
```



---

class: big-text, middle, center, animated, fadeIn

there are *many* ways of<br>modelling time-series data



---

class: big-text, middle, animated, fadeIn

**ARIMA** models combine one or more:

  1. **auto-regression**: *lagged* values of the DV
  2. **integration**: *differenced* values of the DV
  3. **moving-average** values of the DV



---

class: big-text, middle, animated, fadeIn

`arima()` in the `stats` package

`Arima()` in the `forecast` package

`auto.arima()` in the `forecast` package

**`ARIMA()` in the `fable` package**

`stan_sarima()` in the `bayesforecast` package


---

class: middle, center, animated, fadeIn

<span style="font-size: 750%">`r fontawesome::fa("arrow-right", fill = "#0097A9")` &nbsp; `r fontawesome::fa("r-project", fill = "#002855")`</span>



---

class: big-text, middle, animated, fadeIn

# Thanks!

These slides:<br>**lesscrime.info/slides/time-series-models-nscr**

`r fontawesome::fa(name = "envelope")` **matthew.ashby@ucl.ac.uk**

`r fontawesome::fa(name = "twitter", fill = "#1DA1F2")` **@LessCrime**

<p style="font-size: 90%;">Cover photo by <a href="https://unsplash.com/@qrenep?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Rene Böhmer</a> on <a href="https://unsplash.com/s/photos/data?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Unsplash</a></p>
